%!TEX root = slides.tex

\section{Computational complexity}


\begin{frame}{Bubble sort for fixed size input}
  \begin{alertblock}{How many comparisons does bubble sort make for different inputs of the same size?}
    \begin{itemize}
      \item Suppose we have a list with 5 elements.
      \item Best case scenario: list already sorted.
      \item Worst case scenario: list reverse sorted.
    \end{itemize}
    \[ [1,2,3,4,5] \]
    \[ [5,2,1,4,3] \]
    \[ [5,4,3,2,1] \]
  \end{alertblock}
\end{frame}


\begin{frame}{Bubble sort different sized inputs}
  \begin{alertblock}{How many comparisons does bubble sort make for inputs of increasing length?}
    \begin{itemize}
      \item Suppose we have a list with $n$ elements.
      \item Compare the worst case scenario for each value of $n$.
    \end{itemize}
    \[ [3,2,1] \]
    \[ [4,3,2,1] \]
    \[ [5,4,3,2,1] \]
    \[ [6,5,4,3,2,1] \]
  \end{alertblock}
\end{frame}


\begin{frame}{Average vs. worst case}
  \begin{table}
    \begin{tabular}{p{2cm}rr}
      Input & Algorithm A & \hspace{1cm} Algorithm B \\
      \hline
      (1,2,3) &  1ms &  1ms \\
      (1,3,2) &  1ms &  5ms \\
      (2,1,3) &  2ms &  4ms \\
      (2,3,1) &  2ms &  5ms \\
      (3,1,2) &  2ms &  5ms \\
      (3,2,1) & 10ms &  4ms \\
      \hline
      Average &  3ms &  4ms \\
      Worst   & 10ms &  5ms
    \end{tabular}
  \end{table}
  \begin{center}
    Would you choose Algorithm A or Algorithm B?
  \end{center}
\end{frame}


\begin{frame}{Terminology of complexity}
  \begin{alertblock}{How does the number of operations change as $n$ increases?}
    \begin{description}
      \item[Linear] Multiply $n$ by a constant, add a constant. (Special category of polynomial time.)
      \item[Polynomial] Multiply $n$ by itself a fixed number of times, multiply by a constant. (Can add such terms together.)
      \item[Exponential] Raise a constant to the power of $n$. (Rate of change is still exponential.)
      \item[Logarithmic] Opposite of exponential.
    \end{description}
  \end{alertblock}
\end{frame}


\begin{frame}[fragile]{Terminology of complexity (graph)}
\begin{center}
\begin{tikzpicture}
  \begin{axis}[xmin=0, domain=0:7, axis x line=bottom, axis y line=left, legend style={at={(0.5,0.5)},anchor=south}]
    \addplot[red]   {pow(2,x)};
    \addplot[black] {pow(x,2)};
    \addplot[blue]  {2*x};
    \addplot[cyan]  {ln(x)/ln(2))};
    \legend{$2^n$,$n^2$,$2n$,$log(n)$};
  \end{axis}
\end{tikzpicture}
\end{center}
\end{frame}


\begin{frame}{Linear}
  \[ f(n) = a_0 + a_1 n \]
  \begin{alertblock}{How many pairs of shoes does a centipede need?}
    \begin{itemize}
      \item Let's say a centipede has 100 feet.
      \item Then every centipede needs 100 shoes.
      \item That's 50 pairs of shoes.
      \item So 2 centipedes need 100 pairs, 3 need 150 pairs, etc.
      \item So $n$ centipedes need $50n$ pairs of shoes.
      \item Linearity is familiar, and most people's default assumption.
      \item You take the input, multiply by a constant, and add another constant.
    \end{itemize}
  \end{alertblock}
\end{frame}


\begin{frame}{Polynomial}
  \[ f(n) = a_0 + a_1 n + a_2 n^2 + a_3 n^3 + \ldots \]
  \begin{alertblock}{What is the volume of a cube of side $n$?}
    \begin{itemize}
      \item Suppose we have a cube with sides of length 1 metre.
      \item The volume of the cube is $1 \times 1 \times 1 = 1$ metres cubed.
      \item Suppose the cube has sides of length 2 metres instead.
      \item The volume of the cube is $2 \times 2 \times 2 = 8$ metres cubed.
      \item In general, for sides of length $n$, the volume is $n^3$.
    \end{itemize}
  \end{alertblock}
\end{frame}


\begin{frame}{Exponential}
  \[ f(n) = a^n \]
  \begin{alertblock}{How many numbers can we represent with $n$ bits?}
    \begin{itemize}
      \item Consider the case of four bits -- imagine four placeholders \textbf{\fbox{?}}\textbf{\fbox{?}}\textbf{\fbox{?}}\textbf{\fbox{?}}
      \item Each placeholder can contain either 0 or 1.
      \item There are $2 \times 2 \times 2 \times 2 = 2^4 = 16$ different numbers.
      \item Add another bit, how many numbers is it now?
      \item It's $2 \times 2 \times 2 \times 2 \times 2 = 2^5 = 32$.
      \item Generally $n$ bits can represent $2^n$ numbers.
    \end{itemize}
  \end{alertblock}
\end{frame}


\begin{frame}{Logarithmic}
  \[ f(n) = log_a n \]
  \begin{alertblock}{How many bits do we need to represent $n$ numbers?}
    \begin{itemize}
      \item If we have $n$ bits we can represent $2^n$ numbers.
      \item If we want to represent $n$ numbers, how many bits to we need (at a minimum)?
      \item The inverse operation to exponentiation is logarithm.
      \item Remember, $a^n = b$ means $log_a b = n$.
    \end{itemize}
  \end{alertblock}
\end{frame}

\begin{frame}{Big-O (Sipser)}
  \begin{definition}
    Let $f$ and $g$ be functions $f,g: \mathbb{N} \rightarrow \mathbb{R}^+$.
    We say that $f(n) = O(g(n))$, or $f$ is \emph{big-O} of $g$, if positive integers $c$ and $n_0$ exist such that, for every integer $n$ greater than or equal to $n_0$, $f(n) \le cg(n)$.
  \end{definition}
  \begin{alertblock}{Example}
    Let $f$ be the function $f(n) = 5n^3 + 2n^2 + 22n + 6$.
    We'll prove that $f$ is big-O of $n^3$ ($f = O(n^3)$).
    Let $c$ be $6$ and $n_0$ be $10$.
    Is the following true, for all $n$ greater than or equal to 10, $5n^3 + 2n^2 + 22n + 6 \le 6n^3$?
    Note that as $n$ increases ($n=10,n=11,n=12,\ldots$), $f(n)$ also increases.
    Also note that $f(10) = 5426$ and $6g(10) = 6000$.
  \end{alertblock}
  \citeurl{math.mit.edu/~sipser/book.html}
\end{frame}


\begin{frame}[fragile]{Big-O example graph}
\begin{center}
\begin{tikzpicture}
  \begin{axis}[xmin=0, domain=0:30, axis x line=bottom, axis y line=left, legend style={at={(0.4,0.6)},anchor=south}]
    \addplot[red]   {6*pow(x,3)};
    \addplot[black] {5*pow(x,3) + 2*pow(x,2) + 22*x + 6};
    \legend{$6n^3$,$5n^3 + 2n^2 + 22n + 6$};
  \end{axis}
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}[fragile]{Smaller values of $n$}
\begin{center}
\begin{tikzpicture}
  \begin{axis}[xmin=0, domain=0:8, axis x line=bottom, axis y line=left, legend style={at={(0.4,0.6)},anchor=south}]
    \addplot[red]   {6*pow(x,3)};
    \addplot[black] {5*pow(x,3) + 2*pow(x,2) + 22*x + 6};
    \legend{$6n^3$,$5n^3 + 2n^2 + 22n + 6$};
  \end{axis}
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}[fragile]{Bubble sort is $O(n^2)$}
\begin{center}
\begin{tikzpicture}
  \begin{axis}[xmin=0, domain=0:8, axis x line=bottom, axis y line=left, legend style={at={(0.4,0.6)},anchor=south}]
    \addplot[red]   {x^2};
    \addplot[black] {x*(x-1)/2};
    \addplot[cyan]   {x*ln(x)};
    \legend{$n^2$,$n(n-1)/2$,$n\log(n)$};
  \end{axis}
\end{tikzpicture}
\end{center}
\end{frame}


\begin{frame}{Polynomial time}
  \begin{definition}
    An algorithm is said to be solvable in \emph{polynomial time} if the number of steps required to complete the algorithm for a given input is $O(n^k)$ for some nonnegative integer $k$, where $n$ is the complexity of the input.
  \end{definition}
  \vspace{0.5cm}
  \metroset{block=fill}
  \begin{block}{$P$ complexity class}
    The $P$ complexity class is the set of problems for which there exists, for each such problem, at least one algorithm to solve that problem in polynomial time.
  \end{block}

  \citeurl{mathworld.wolfram.com/P-Problem.html}
\end{frame}


\begin{frame}{Polynomial time on a Turing machine}
  \begin{description}
    \item[Sorting] algorithms are usually compared in terms of comparisons.
    \item[Other algorithms] might be compared in terms of something else, like iterations.
    \item[With Turing machines] we can use the number of times we look up the state table.
    \item[The size] of the input can be the length of the input on the tape initially.
  \end{description}
\end{frame}


\begin{frame}{Recap on Languages}
  \begin{description}
    \item[Alphabet] Finite set of symbols, denoted $\Sigma$.
    \item[String] Sequence of symbols, $w$ from $\Sigma$.
    \item[Language] Set of strings, denoted $L$.
    \item[Length] Of a string, denoted $|w|$  .
    \item[Empty string] Unique string of length 0, denoted $\epsilon$.
  \end{description}
\end{frame}


\begin{frame}{Kleene star}
  \begin{description}
    \item[Word concatenation:] $w_1 w_2$ is the concatenation of strings $w_1$ and $w_2$.
    \item[String concatenation:] $L_1 L_2$ is the language resulting from the concatenation of all strings in $L_1$ and all words in $L_2$, in that order.
    \item[Powers:] $L^0 = \{ \lambda \}$, $L^1 = L$ and $L^{n+1} = L^n L$ for all $n > 1$.
  \end{description}
  \vspace{0.5cm}
  \begin{block}{Kleene Star}
     \[ L^* =  \bigcup_{i=0}^{\infty} L^i \]
  \end{block}
  Note that treating the alphabet $\Sigma$ as a language in itself, we get that $\Sigma^*$ is the set of all words over $\Sigma$.
\end{frame}


\begin{frame}{Example}
  \begin{description}
    \item[$\Sigma$] $\{ 0, 1 \}$
    \item[$L$] $\{ 00, 01, 10, 11 \}$
    \item[$w_1$] $01$
    \item[$w_3$] $11$
    \item[$w_1 w_3$] $0111$
    \item[$\Sigma^*$] $\{ \lambda, 0, 1, 00, 01, 10, 11, 001, 010, \ldots \}$
    \item[$L^*$] $\{ \lambda, 00, 01, 10, 11, 0000, 0001, \ldots \}$
    \item[$L^+$] $\{ 00, 01, 10, 11, 0000, 0001, \ldots \}$
  \end{description}
\end{frame}


\begin{frame}{Decision problems}
  \begin{description}
    \item[Decision problems] are problems where the answer is 0 or 1.
    \vspace{0.5cm}
    \item[Restricting] ourselves to decision problems is convenient and fair.
    \vspace{0.5cm}
    \item[$f:\{0,1\}^n \rightarrow \{0,1\}$] is useful notation for considering decision problems.
    \vspace{0.5cm}
    \item[Turing machines] model decision problems by deciding languages.
  \end{description}
\end{frame}


\begin{frame}{PRIMES}
  \begin{definition}
    $$ PRIMES = \{ 2, 3, 5, 7, 11, 13, ... \} $$
  \end{definition}
  \begin{description}
    \item[PRIMES] is the set of all primes numbers.
    \item[Can] a Turing Machine be designed to decide PRIMES? (Yes)
    \item[Some] people say PRIMES is the decision problem for the set of primes.
    \item[Can] it do it in polynomial time? (Yes)
    \item[Is] PRIMES in P? (Yes, 2002)
  \end{description}
  \citeurl{www.cse.iitk.ac.in/users/manindra/algebra/primality\_v6.pdf}
\end{frame}

\begin{frame}{Modern cryptography}
  \begin{alertblock}{Modern asymmetric key crytography is based on prime numbers}
    It depends on two facts:
    \begin{itemize}
      \item It's easy to find primes (P).
      \item It's hard to decompose a composite number into primes (Not known to be P).
    \end{itemize}
  \end{alertblock}
\end{frame}

\begin{frame}[fragile]{Brute force prime checking}
  \begin{alertblock}{Is n a prime?}
    \begin{minted}{c}
int is_prime(int n) {
  for (int i = 0; i < n; i++) {
    if (n % i == 0)
      return 0;
  }
  return 1;
}
    \end{minted}
  \end{alertblock}
\end{frame}

\begin{frame}[fragile]{More efficient prime checking}
  \begin{itemize}
    \item Suppose $a \times b = n$.
    \item Then $a < b$, $a > b$ or $a = b$.
    \item No matter what, $a \le \sqrt{n}$ and/or $b \le \sqrt{n}$.
    \item So only loop to $\sqrt{n}$.
    \item This still isn't that efficient.
  \end{itemize}
\end{frame}